---
title: "README"
author: "James Durant"
date: "October 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About
This project is intended to serve as a repository of teaching lessons that are written in R markdown that can be render to
different formats (html5, pdf/beamer, MS Word) as well as produce student handout for either the instructor or the student. 

# Usage:

## Lessons
Lessons are stored under *Rmd_Lessons* folder. These are bare Rmd files (without YAML header) for the lessons. The lessons will have an $<$article$>$ and $<$\\article$>$ tag to indicate material that will only appear in the student handout document. 

## Rmd Templates
Rmd templates are stored under _Rmd_templates_ directory. These files are basically the YAML section of the RMD document are combined to the Rmd_lessons to the document to be knit into final format.

## Usage
Under the R directory, there is a function called `createLessonMaster.R` and `RmdFlip.R` which will need to be sourced into your R console will create the lesson:

```{r eval=TRUE}

source("./R/createLessonMaster.R")
source("./R/RmdFlip.R")
```

To use the functions:
```{r eval=FALSE}
createLessonMaster <- function(RmdFiles, type=c("article","beamer","tangohtml", "shower"),
                               headFile = "./Rmd_templates/beamer.Rmd")
```

Where

- RmdFiles is a list of Rmd_lessons (relative path included) of files for the lesson.
- type is the type of document to produce (article is student handout)
- headfile is the Rmd_template to use.

So for example, to create a "shower" html presentation with all the lessons in the *Rmd_lessons*:


```{r eval=FALSE}
x <- list.files("Rmd_lessons", pattern = "[.]Rmd", full.names = TRUE)

createLessonMaster(x, type="shower", headFile = "./Rmd_templates/shower.Rmd")
render("Master_shower.Rmd")
```

The function createLessonMaster will call RmdFlip.R to remove the material between the $<$article$>$ and the $<$\\article$>$ taglines for the standard presenation, and will just remove the tags themselves for the student handout.  


## Introduction to R and RUG talk

To create the RUG talk:
```{r eval = TRUE, message=FALSE, results='hide', warning=FALSE}
library(knitr)
library(rmarkdown)
x <- list.files("Rmd_lessons", pattern = "[.]Rmd", full.names = TRUE)
x <- grep("getPacks|Background|help", x, value=TRUE)
createLessonMaster(x, type="shower", headFile = "./Rmd_templates/shower.Rmd")
render("Master_shower.Rmd")
```


