# Data Import

## Excel Data
 
The _readxl_ package provides functions for importing Excel Spreadsheets:
- Data exported to Excel from CDC's Food Outbreak Online Database (FOOD) (https://wwwn.cdc.gov/foodborneoutbreaks/)
```{r read_excel}
library(readxl)
fddf <- read_excel("../data/FoodData.xlsx")
```

- in addition to local files, you can also use http, https, and ftp locations.

## Comma Separated Values

Data from https://www.cdc.gov/growthcharts/percentile_data_files.htm
```{r read_csv}
lms_chart <- read.csv("https://www.cdc.gov/growthcharts/data/zscore/wtageinf.csv")
```

## Tab delimited data
- example from MMWR download from CDC's Wonder.
- data are downloaded as plain text, with tabular data in lines 22 - 66.
- `read.delim` with arguments skip and nrows to identify data
```{r download mmwr}
mmwr2017_31 <- read.delim("https://wonder.cdc.gov/mmwr/mmwr_reps.asp?mmwr_year=2017&mmwr_week=31&mmwr_table=2M&request=Export&mmwr_location=", skip =21, nrows = 66, na.strings = "-", header = FALSE)

names(mmwr2017_31) <- c("Reporting Area",
"Salmonellosis current week",
"Salmonellosis previous 52 weeks median", 
"Salmonellosis previous 52 weeks maximum", 
"Salmonellosis cummulative for 2017",
"Salmonellosis cummulative for 2016",
"STEC current week",
"STEC previous 52 weeks median", 
"STEC previous 52 weeks maximum", 
"STEC cummulative for 2017",
"STEC cummulative for 2016",
"Shigellosis current week",
"Shigellosis previous 52 weeks median", 
"Shigellosis previous 52 weeks maximum", 
"Shigellosis cummulative for 2017",
"Shigellosis cummulative for 2016")
```

## SAS Export Files
To download and load NHANES 2013 - 2014 demographic files
```{r NHANES}
library(foreign)
demo_h <- read.xport("../data/demo_h.xpt")
```


## SAS bat files
Data from ATSDR's National Toxic Substances Incident Program (https://www.atsdr.cdc.gov/ntsip/index.html)
```{r NTIS}
library(haven)
ntis_2014 <- read_sas("https://www.atsdr.cdc.gov/ntsip/docs/ntsippublic2014.sas7bdat")
```

## Exercise
- Import the Salmonellosis data from Excel into a data.frame or Tibble in R.
- Import the Friendship AGE Surveillance Log into a data.frame or Tibble in R.
- Change Date to include year (2015)
- Convert to Date format using `as.Date`





```{r exercise_import}
library(readxl)
library(dplyr)
salmonella <- read_excel("../data/Salmonellosis.xlsx")

cruise <- read_excel("../data/Friendship AGE Surveillance Log.xls", skip = 7) %>%
  select(Date, Time, Age = X__1, Sex = X__2, PsCrw = X__3, Cabin = X__4, Meal_seat = X__5, 
         Pos = X__6, Date_rep = Date__1, Time_rep = Time__1, Diarrhea_eps = `#`, 
         Blood = `Blood Y/N`, Vomiting_eps = `#__1`, Fever = `Y/N`, 
         Temp =  ends_with("F"), Cramps = `Y/N__1`,
         Headache = `Y/N__2`, Myalgia = `Y/N__3`, Spec_req = `Y/N__4`,
         Spec_rec = `Y/N__5`, Antidiarrheal = `Y/N__6`, Reportable = `Y/N__7`,
         Underlying.Illness = `X__7`
         )

cruise$Date <- paste(cruise$Date, "2015", sep="/")
cruise$Date <- as.Date(cruise$Date, format = "%m/%d/%Y")

```


# Analysis and Visualization




## NTIS 2014

### Barplot 
Let's say you want to find out what the number of incidents are per state. The `table` function would be useful:

```{r stateNTIS}
ntis_state <- table(ntis_2014$STATE)
barplot(ntis_state, main = "Number of NTIS Incidents", col="darkblue")
ntis_state
```

### Exercise
Create a table of Chemicals that have had more than 1evacuations ordered (bonus create a barplot):

```{r stateNTIS_evac}
evac_ordered <- ntis_2014[ntis_2014$EVAC_ORD == "Y",]
ntis_evac <- table(evac_ordered$CHEM1)
ntis_evac <- ntis_evac[order(ntis_evac, decreasing = TRUE)]

par(mar=c(5,15,4,2))
barplot(ntis_evac[ntis_evac >1], main = "Number of NTIS Evaculation Incidents, 2014", col="darkblue", las=1, cex.names = 0.75, horiz = TRUE)
```


## Histograms
- How many people lived within a quarter mile of the release?
- To make a histogram of the data, you can use the function `hist`

```{r}
hist(ntis_2014$LIVEQTR, main = "Number of People Inside 1/4 Mile of Release",
     xlab = "Number of people", col="red")
```

# MMWR Data

## Plot Rates Relative to Median Weekly Rate

We will for this exercise identify reporting areas that have salmonella cases 
higher than median over past 52 weeks.

```{r mmwr2017_salmonella}

sal_mmwr <- subset(mmwr2017_31, !is.na(`Salmonellosis current week`))
sal_mmwr <- sal_mmwr[sal_mmwr$`Salmonellosis current week` > 
           sal_mmwr$`Salmonellosis previous 52 weeks median`,]

sal_mmwr$`Reporting Area` <- factor(sal_mmwr$`Reporting Area`)
sal_mmwr$`Salmonellosis previous 52 weeks maximum` <- as.numeric(
  as.character(
    sal_mmwr$`Salmonellosis previous 52 weeks maximum`
      )
  )
  
par(mar=c(10,4,4,4))
bp_sal <- barplot(sal_mmwr$`Salmonellosis current`[order(sal_mmwr$`Salmonellosis current`)],
        names = sal_mmwr$`Reporting Area`[order(sal_mmwr$`Salmonellosis current`)],
        las=2)

points(x = bp_sal, y = sal_mmwr$`Salmonellosis previous 52 weeks median`[order(sal_mmwr$`Salmonellosis current`)], pch =19)

points(x = bp_sal, y = sal_mmwr$`Salmonellosis previous 52 weeks maximum`[order(sal_mmwr$`Salmonellosis current`)], pch =17)
legend("topleft", pch = c(19,17), legend = c("52 wk. Median", "52 wk. Maximum"))
```

## xy Plots

```{r}
mmwr2017_31$`Salmonellosis cummulative for 2017` <- as.numeric(as.character(mmwr2017_31$`Salmonellosis cummulative for 2017`))

mmwr2017_31$`Salmonellosis cummulative for 2016` <- as.numeric(as.character(mmwr2017_31$`Salmonellosis cummulative for 2016`))

plot(`Salmonellosis cummulative for 2017` ~ `Salmonellosis cummulative for 2016`,
     data = mmwr2017_31, pch =16)
```

## Correlation and Linear Models

To see if there is a correlation, you can use either `cor()` or `lm()` functions.

```{r}
cor(mmwr2017_31$`Salmonellosis cummulative for 2017`, 
    mmwr2017_31$`Salmonellosis cummulative for 2016`, use = "complete.obs")

cor(mmwr2017_31$`Salmonellosis cummulative for 2017`, 
    mmwr2017_31$`Salmonellosis cummulative for 2016`, use = "complete.obs",
    method="kendall")

salmon.lm1 <- lm(
  `Salmonellosis cummulative for 2017` ~ `Salmonellosis cummulative for 2016`,
     data = mmwr2017_31)

summary(salmon.lm1)
```
```{r}
plot(`Salmonellosis cummulative for 2017` ~ `Salmonellosis cummulative for 2016`,
     data = mmwr2017_31, pch =16)
lines(mmwr2017_31$`Salmonellosis cummulative for 2016`, predict(salmon.lm1, newdata = mmwr2017_31))
```


# Cruise Ship Data

## Draw and Epi-Curve


```{r}
library(EpiCurve)
EpiCurve(cruise, date = "Date", period="day")
```

## Recoding a Variable

The meal seat variable is a little difficult to interpret:

```{r}
table(cruise$Meal_seat)
```


- We want to recode into "Crew", "My Time", "Romeo & Juliet", "Other"  
- grep uses Regular Expressions to select rows with matching strings.

```{r}
cruise$Meal_seat[grep("Crew|Staff", 
                      cruise$Meal_seat, 
                      ignore.case = TRUE)] <- "Crew"
cruise$Meal_seat[grep("My ?time", 
                      cruise$Meal_seat, 
                      ignore.case = TRUE)] <- "My Time"
cruise$Meal_seat[grep("Romeo \\& Juliet|R\\&J", 
                      cruise$Meal_seat, 
                      ignore.case = TRUE)] <- "Romeo & Juliet"

cruise$Meal_seat[grep("Crew|My Time|Romeo \\& Juliet", 
                      cruise$Meal_seat, invert=TRUE)] <- "Other"
table(cruise$Meal_seat)
```

To graph our recoded variables in the Epi-Curve function:
```{r}
# we will use nceh colors
nceh_cols <- c(rgb(0,130/255,101/255), rgb(0,89/255,132/255), rgb(71/255,103/255, 81/255), rgb(84/255, 25/255,0))
               
EpiCurve(cruise, date="Date", period = "day", cutvar = "Meal_seat", colors = nceh_cols)
```

## Boxplots
Boxplots are useful summaries for 


